Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-09-02 doc/response_to_reviewer_R1.Rmd

•	It is always helpful to know the study site. I recommend including some pictures of the forest and study area.

RESPONSE:

•	The methodology should clearly state whether the 30 species selected for the study represent the functional groups. It should include the number of species per Deciduousness group.

**RESPONSE: Thanks for this comment. We agree that it is useful to include the functional representativeness of the species. As we mention in the methods, our metric of deciduousness is a continuous variable and not categorical. And so, reporting the number per deciduousness group would require arbitrarily discretising this continuous variable. Instead, we now report a supplementary figure showing the distribution of deciduousness among the selected species.**

@@

239-242: I think that this extra information is unnecessary, since although this measurement is interesting, it was not the one applied in this study.

**RESPONSE: We have now removed this statement regarding the alternate measure of deciduousness that was considered.**

301-307: The figure that supports this paragraph needs to be cited. I would think it would be figure 2a, but the species Gluta obovata is not there. I did not see the analysis by dominant species anymore, if the study does not focus on these species, I recommend removing it from Figure 2 since it confuses the reader.

**RESPONSE: Thanks for this comment. We have revised Figure 2 to remove species lines. Originally, we had chosen to include the top 10 species in the plot to illustrate both synchrony and divergence in species responses, but the results text reported on growth and sensitivity values across all species considered (n=30), shown in the Supplementary Appendix. We agree that this misalignment in species choice was confusing and have removed it from Figure 2 and referenced the relevant figure in the results text.**

@@

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

***RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity. In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees. TO CLARIFY IN TEXT***

@@

RESPONSE:

******************************************
Editor's comments to the author(s):

Editor
Editors Comments for the Author(s):
Two reviewers have reviewed this study and see a lot of value and importance in the data and conclusions. They provide valuable comments that should be able to be addressed in a major revision.

**RESPONSE: Thanks for your favourable review of the manuscript. We have now addressed all the reviewer comments, which have improved the manuscript. We hope that the manuscript can be considered for publication in Ecology Letters.**

### 2025-09-02 doc/response_to_reviewer_R1.Rmd

Referee: 1

Comments for the Authors
General comment

The response of tropical forests to drought is a central topic in global ecology and climate change, directly aligned with the interests of Ecology Letters. The novel of this work is the central argument that drought sensitivity is not a fixed species trait, but rather the result of the interaction between drought characteristics, adaptations, and the microenvironment, provides conceptual originality. The authors used a 14-year series of dendrometer bands in a big plot with 1,820 trees of 30 species, which is a high-value dataset. In addition, the statistical analysis based on Bayesian models and a causal framework to disentangle trait, size, and microenvironment effects appears to be a sufficiently robust approach to address predictions. However, I have some concerns regarding the with the current version that should be addressed by the authors, which are appointed below.

**RESPONSE: Thank you for your favourable review of the manuscript and for the constructive comments. We have now revised the manuscript based on these suggestions and that of the other reviewer which have helped improve the mansucript. Our detailed comments are below.**

@@

55: Could you briefly explain the three climatically distinct droughts?

RESPONSE:

162-163: Until when was the measurement? It is important to add the last year of measurement.

**RESPONSE: Our analysis includes measurements until January 2023, making 2022 the last growing year considered. We have now added this detail. It now reads:  
*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***

@@

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

***RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity. In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees. TO CLARIFY IN TEXT***

Table 1: “other size effect” is DBH?

**RESPONSE: Yes, other size effect is DBH, accounting for crown exposure. We have now made that clear.**

@@

Figure 1a: “Dy” is not correct

**RESPONSE: Thanks for noticing this error. We have fixed this to say "Day of year".**

Fig. 2. It is unnecessary to include the years 2023 and 2024 on the y-axis, as they were not evaluated. I think this figure could be analyzed further. For example, it took two years after 2014 for growth to recover to previous year (for ALL species) unlike in 2010 and 2020, when it rebounded within a year to levels equal to or higher than the previous year.

*RESPONSE: Thanks for this comment. We have removed 2023 and 2024 from the x-axis. While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates. The pattern that you describe for the 2015 drought is interesting, but this is primarily because drought conditions were prolonged into 2016 - this was a multi-year drought - and the ENSO conditions switched only late in 2016, leading to a late monsoon. Our analysis does not have the power to analyse multi-year droughts separately from single year droughts, and therefore we chose not to analyse and interpret these cycles.*

@@

Referee: 2

Comments for the Authors
This study investigated the impact of three different drought events on tropical tree radial growth based on the band dendrometer monitoring of 50 ha tropical seasonally dry plot in western Thailand. The study found that wet season drought in 2015 had the most negative impact on the tree growth compared with the other two drought events. In addition, the results also showed that the impact of deciduousness, soil availability and canopy exposure had significant impact on the growth response to drought. The present study is useful to understand the impact of drought events. The data is very strong and the manuscript is well written, the analysis were also well done. The present study is helpful to understand the impact of drought events on the growth of tropical trees. I had some minor comments hope can help to improve the manuscript.

**RESPONSE: Thank you for your favourable review of our manuscript and for the constructive comments. This revision following comments has improved the clarity and interpretability of our manuscript. Our detailed responses to your comments are given below.**

@@

1 Line 175 “Using the late wet/early dry dendrometer…” Since there were two time of censuses (January and June) for the measurement, can the author clarify which time do they use to calculate the annual growth? Based on the measurement of January or June?

**RESPONSE: The January measurement was used to calculate annual growth. We have now included this detail at in now reads:  
*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***

@@

2 The topographic wetness index. Are there any measured soil moisture data in this plot? Since this index is only based on topography, it will be good if the author could provide some measured soil moisture data to calibrate the prediction based on model prediction.

RESPONSE:

3 In the case of deciduousness, the authors indicated that they used the deciduous index from Williams et al. (2008). Since different species might shed leaves in different season. Did the authors consider the impact of the timing of deciduousness on the growth?

*RESPONSE: Thanks for this interesting comment. The timing of deciduousness may be an important factor to consider, however, from the data reported in Williams et al. (2008), leaf flush for most species is reported during the month after the spring equinox in late March, independent of the duration of leaflessness. Moreover, the timing of deciduousness can differ from year to year, and would require direct measurement at the tree to answer this interesting question.*

### 2025-09-03 doc/response_to_reviewer_R1.Rmd

•	Predictions (i–iii) are clearly formulated, but the Results section is at times overly dense with statistical detail. It would be helpful to present more explicitly which findings support each hypothesis.

RESPONSE:

•	276: This section is important, but it's not a relevant result; rather, it's a methodological aspect that allowed the study to be conducted. I think it should be included in the materials and methods section.

**RESPONSE: Thanks for this suggestion. We have now moved the description of the droughts to the methods section and have only retained a brief description of the differences among droughts in the results.**

@@

55: Could you briefly explain the three climatically distinct droughts?

**RESPONSE: The three climatically distinct droughts**

162-163: Until when was the measurement? It is important to add the last year of measurement.

**RESPONSE: Our analysis includes measurements until January 2023, making 2022 the last growing year considered. We have now added this detail. It now reads:  
"Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022."**

@@

Comments for the Authors
This study investigated the impact of three different drought events on tropical tree radial growth based on the band dendrometer monitoring of 50 ha tropical seasonally dry plot in western Thailand. The study found that wet season drought in 2015 had the most negative impact on the tree growth compared with the other two drought events. In addition, the results also showed that the impact of deciduousness, soil availability and canopy exposure had significant impact on the growth response to drought. The present study is useful to understand the impact of drought events. The data is very strong and the manuscript is well written, the analysis were also well done. The present study is helpful to understand the impact of drought events on the growth of tropical trees. I had some minor comments hope can help to improve the manuscript.

**RESPONSE: Thank you for your favourable review of our manuscript and for the constructive comments. This revision following comments has improved the clarity and interpretability of our manuscript. Our detailed responses to your comments are given below.**

1 Line 175 “Using the late wet/early dry dendrometer…” Since there were two time of censuses (January and June) for the measurement, can the author clarify which time do they use to calculate the annual growth? Based on the measurement of January or June?

**RESPONSE: The January measurement was used to calculate annual growth. We have now included this detail at in now reads:  
"Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022."**

@@

2 The topographic wetness index. Are there any measured soil moisture data in this plot? Since this index is only based on topography, it will be good if the author could provide some measured soil moisture data to calibrate the prediction based on model prediction.

*RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available. HKK, being fairly remote, has limited access to power sources and other infrastructure that are needed to maintain such data collection. We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.*

@@

Fig S7 “…included in the analyses in the 2010 and 2015 datasets”, do you mean “2010, 2015 and 2020”?

**RESPONSE: Thanks for noticing this error. We have changed this to "2010, 2015 and 2020".**

Fig. S8 …need to add X and Y axes for the figures.

**RESPONSE: Thanks for noticing this. We have made the change.**

### 2025-09-03 doc/manuscript_hkk_drought_sensitivity.Rmd
updated summary table and some draft responses
some more comments

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses). -->
We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census for each tree $(d_{i, t})$ using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
We used a direct measurement of $dia_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($\Delta day$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{\Delta day}{365}$).

@@

@@

## Drought characteristics {-}

@@

## Growth responses to drought {-}

@@

Median annual increment across yeaes varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.

>>>>

Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022. 
<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses). -->
We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census for each tree $(d_{i, t})$ using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
We used a direct measurement of $dia_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($\Delta day$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{\Delta day}{365}$).

@@

<!-- ## Drought characteristics {-} -->

@@

@@

## Tree growth responses to drought {-}

@@

Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
